---
title: "Project1"
author: "Sean Fahey"
date: "2023-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(ggfortify)
library(plotly)

pdata <- read.csv("pcosdata.csv")

```

# Background
Polycystic ovary syndrome (PCOS) is a complex condition characterized by elevated androgen levels, menstrual irregularities, and/or small cysts on one or both ovaries.

The disorder can be morphological (polycystic ovaries) or predominantly biochemical (hyperandrogenemia). 

Hyperandrogenism, a clinical hallmark of PCOS, can cause inhibition of follicular development, microcysts in the ovaries, anovulation, and menstrual changes.

```{r inspect dataset}
pdata$Oligomenorrhea <- as.factor(pdata$Oligomenorrhea)
pdata$Polycystic.Ovarian.Morphology <- as.factor(pdata$Polycystic.Ovarian.Morphology)
pdata$Hyperandrogenism <- as.factor(pdata$Hyperandrogenism)
pdata$Category <- as.factor(pdata$Category)

pdata <- mutate(pdata, Weight.Category = ifelse((BMI > 25), "Overweight", "Healthy/Low"))
pdata$Weight.Category <- as.factor(pdata$Weight.Category)

head(pdata)
names(pdata) #verify all variables are present
summary(pdata$BMI)
hist(pdata$BMI, breaks=15)
```
# Hypothesis
Increasing testosterone and age are more likely to result in an elevated BMI. 

```{r initial plots to explore dataset}
ggplot(
  pdata,
  aes(
    x=Testosterone.ng.mL,
    y=BMI
    )
  ) + 
  geom_point(
    size=1) + 
  theme_minimal() +
  geom_smooth(
    formula = 'y ~ x',
    method = 'lm',
    se=FALSE) +
  labs(
    x="Testosterone (ng/mL)",
    y="BMI",
    title="Testosterone Levels v. BMI"
  )
ggsave("Testosterone-v-BMI.png", height=4.5, width = 4.5, units = "in")


ggplot(
  pdata,
  aes(
    x=Age,
    y=BMI,
    )
  ) + 
  geom_point(
    size=1) + 
  theme_minimal() +
  geom_smooth(
    formula = 'y ~ x',
    method = 'lm',
    se=FALSE) +
  labs(
    x="Age",
    y="BMI",
    title="Participant Age v. BMI"
  )
ggsave("Age-v-BMI.png", height=4.5, width = 4.5, units = "in")


ggplot(
  pdata,
  aes(
    x=Testosterone.ng.mL,
    y=BMI,
    color=Age
    )
  ) + 
  scale_colour_gradient(
  low = "red",
  high = "green",
  space = "Lab",
  na.value = "grey50",
  guide = "colourbar",
  aesthetics = "colour"
  ) +
  geom_point(
    size=1,
  ) + 
  theme_gray() +
  geom_smooth(
    formula = 'y ~ x',
    method = 'lm',
    se=FALSE) +
  labs(
    x="Testosterone (ng/mL)",
    y="BMI",
    title="Testosterone Levels v. BMI",
    subtitle="colored by age"
  )
ggsave("Testosterone+Age-v-BMI.png", height=4.5, width = 6, units = "in")

```

There is a slight positive correlation between Age/BMI and between Testosterone/BMI. However, the combination of Age and Testosterone don't seem to have a clear interaction.

```{r setting up linear model}
m1 = glm(BMI ~ Testosterone.ng.mL + Age, data=pdata)
```

```{r inspecting lienar model}
summary(m1)
anova(m1, test="F")
autoplot(m1)

```

While there is a statistically significant correlation between both Testosterone (t-val: 2.442, Df: 174, p-val: 0.01562), and Age (t-val: 3.240, Df: 174, p-val:0.00144), the residual deviance is much higher than the number of degrees of freedom. To correct for this, we will run a poisson model instead.

```{r setting up possion GLM model, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

m2 = glm(BMI ~ Testosterone.ng.mL + Age, data=pdata, family="poisson")
```

```{r inspecting possion GLM model}
summary(m2)
anova(m2, test="F")
autoplot(m2)
```

There is a significant positive correlation between both Testosterone and BMI (z-val: 2.637, Df: 174, p-val: 0.008374) and between Age and BMI (z-val: 3.497, Df: 174, p-val: 0.000471). Also the residual deviance is much lower than the initial model, so the change improved our analysis.

# Final plot

```{r final figure supporting the GLM model}
fig <- plot_ly(
  pdata, 
  x = ~Age, 
  z = ~BMI, 
  y = ~Testosterone.ng.mL, 
  color = ~BMI,
  marker = list(
    size = 5,
    line = list(
      color = 'black',
      width = 2
    )))
fig

htmlwidgets::saveWidget(as_widget(fig), "3D-plot.html")
```

