---
title: "Homework 10a"
author: "Sean Fahey"
date: "2023-03-20"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(car)
library(psych)
library(plotly)
```



```{r import data}
gut = read.csv("gut_diversity.csv", header=TRUE)

gut$age = scale(gut$age)
gut$exercise = scale(gut$exercise)
gut$weight = scale(gut$weight)

gut$subject = as.factor(gut$subject)
summary(gut)
```

## In Class Work

Ancova is anova plus regression analysis.

Fit a linear model using lm(), with both IVs but no interaction. Look at the 
intercept and slopes.

```{r 2 IVs with no interactions}
model.both_iv = lm(diversity ~ age + exercise, data=gut)
summary(model.both_iv)
```

Fit a separate simple linear regression model using lm() for each IV. Compare 
the intercept and slopes to the multiple regression model.

```{r 2 IVs assessed independantly}
model.age_only = lm(diversity ~ age, data=gut)
model.exercise_only = lm(diversity ~ exercise, data=gut)

summary(model.age_only)
summary(model.exercise_only)
```

Intercept is same for both. Slope changes.


Simple linear regression of diversity (DV) on weight (IV). Look at 
significance, either with summary() or anova()
```{r weight assessed independantly}
model.weight_only = lm(diversity ~ weight, data=gut)
summary(model.weight_only)
```

Multiple linear regression of diversity (DV) on exercise (IV1) and weight 
(IV2), no interaction. Look at significance of weight now. 
```{r exercise and weight, no interaction}
model.ex_wt = lm(diversity ~ weight + exercise, data = gut)
summary(model.ex_wt)
```
Adding exercise changed the contribution of weight significantly. Looks like
exercise complicated things a bit.


Plot weight v. exercise to see what's going on.
```{r plot weight v. exercise}

ggplot(
  data=gut,
  aes(
    x = weight,
    y = exercise
    )
  ) + 
  geom_point()

cor(gut$weight, gut$exercise)
```
Looks like there's a correlation between the 2 IVs. That explains what we're
seeing with the diminished effect of weight.

One option: just drop one of the variables.
Another option: model the interaction

```{r quick distraction on plotting}
# pairs just shows a plot of each variable against each other. Sort of
# nice for quick visualization.
pairs(~age+exercise+weight, data=gut)

# pairs.panels (relies on psych package, can't handle ~ inputs) is also cool!
pairs.panels(gut[ ,3:5])

```


## Homework 10a

Fit the gut OTU Shannon diversity model with an interaction (DV = diversity, 
IV = age, exercise, and their interaction).
```{r question 1, fit model with interaction}
model.age.ex = lm(diversity ~ age * exercise, data=gut)

```


Test the IVs. There is an interaction, so type III Anova() would probably be 
best so age * exercise doesn't give different results than exercise * age… 
But good news, if ALL your IVs are continuous, you don't have to worry about 
changing the contrasts statement.
  - type III Anova() and summary() give same p-values
  - anova() is still order-dependent unless perfectly balanced (very unlikely 
  with continuous IVs)

```{r question 2, test the IVs}
Anova(model.age.ex, type=3)
summary(model.age.ex)
```

I know the interaction is not significant, but pretend that it is. How would 
you describe, in words (not numbers) the OTU diversity relationship with age 
and exercise? Include your description in the text of your Rmd file.

Question 3:
Gut microbe diversity has a string, positive correlation with exercise. There
is no statistically significant correlation between age and microbe diversity
or between microbe diversity and the interaction between age and exercise.


It is hard to plot when you have >1 continuous IV. Try to come up with a way 
to do this. 

```{r question 4, r plotly 3D scatter plot}
# reloading the gut data set because sclae is somehow breaking it for plotly.
gut2 = read.csv("gut_diversity.csv", header=TRUE)

fig <- plot_ly(
  type="scatter3d",
  mode="markers",
  gut2,
  x = ~exercise, 
  y = ~weight, 
  z = ~diversity,
  color = ~age
  )

fig
```


