---
title: "HW12"
author: "Sean Fahey"
date: "2023-04-03"
output: html_document
---

```{r setup}
library(ggplot2)

```

# In-Class
```{r get dataset}
chimp = read.csv("ChimpBrains.csv", header=TRUE)
chimp$chimpName = as.factor(chimp$chimpName)
chimp$sex = as.factor(chimp$sex)

ragweed = read.csv("Ragweed.csv", header=TRUE)
ragweed$ID = as.factor(ragweed$ID)
ragweed$Site = as.factor(ragweed$Site)

snail = read.csv("SnailGrowth.csv", header=TRUE)
snail$Temperature = as.factor(snail$Temperature)
snail$Crab = as.factor(snail$Crab)

summary(ragweed)
```

### Permutation analysis

In ragweed data set we have 2 groups and we want to decide if they're 
different; "roadcut" and "protected" (Site). 

Step 1, decide test statistic:

- |mean(Roadcut) - mean(Protected)| works well. We get 3.6 here.
  
Step 2, generate a null distribution:

- Figure out what it would look like if site type didn't matter.
- We expect Null distribution to cluster around 0, but not be exactly 0.
- Randomly re-sample/shuffle the data into "roadcut" and "protected" (just 
shuffle the Biomass column basically). Calculate the difference using the 
formula from above.
- Re-do the shuffle, get another value. Do this until we have 10,000 values.
- This tells us what to expect if only sampling error causes the difference
between groups.

```{r calculate observed means}
ragmean = aggregate(Biomass~Site, data=ragweed, FUN = mean)
ragdiff = abs(ragmean$Biomass[1] - ragmean$Biomass[2])

print(ragmean)
print(ragdiff)
```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



