---
title: "Lecture 13"
author: "Sean Fahey"
date: "2023-04-05"
output: html_document
---

```{r setup}
library(ggplot2)
library(Hotelling) # for hotelling's test
library(car)
library(mvnTest)
library(vegan)
```

# In Class
### bottle and iris datasets

univariate: one response
multivariate: more than one response variable

- multiple measurements from the same individual that are **NOT independent**
- can be cont. or cat.
- will use multivariate for final

```{r get bottle dataset ready}
data(bottle.df)
summary(bottle.df)

cat('\n\n')

# get just rows that match bottle 1 & 2
bottle.1 = bottle.df[bottle.df$Number == 1, -1] # leaving off the first column
bottle.2 = bottle.df[bottle.df$Number == 2, -1]

# get means for each column (for each bottle)
colMeans(bottle.1)
colMeans(bottle.2)
```


```{r simple multivariate analysis}
bottle_hotelling_test = hotelling.test(
                            x = bottle.1, 
                            y = bottle.2
                          )

bottle_hotelling_test
```
p-value of 0.002949. Pretty low. Reject the null hypothesis that the elements
in these bottle types is the same.

Use *MANOVA* to compare 2 or more multivariate samples. Like ANOVA, but for
multivariate.

- Compares centers of each group. Each group is a cloud of points on 
n-dimensions (number of response variables). 
- MANOVA only cares about center 
of the group cloud, like how ANOVA cares about the mean for each group's 
response. 

```{r example to visualize multiple response variables}
scatterplotMatrix( # from car package
  ~ Sepal.Length + 
    Sepal.Width + 
    Petal.Length 
  | Species, 
  data=iris
  )

```
Another way to visualize is to just make boxplots for each response for each
species.

```{r MANOVA example}
# using 2 response varaibles to start
manova.1 = manova(
  cbind(Sepal.Length, Sepal.Width) ~ # cbind because we need to do it on both
    Species, 
  data=iris
  )

summary(manova.1)
cat('\n\n\n')
summary.aov(manova.1) # gives p value for each response variable
```
Looks like there's a significant difference between the three species with
respect to the size and shape of the sepal.

Also from the summary.aov, it looks like both response variables matter.

```{r MANOVA for all 4 traits}
manova.2 = manova(
  cbind(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) ~ 
    Species, 
  data=iris
  )

summary(manova.2)
cat('\n\n\n')
summary.aov(manova.2)
```
All 4 response variables contribute to the difference.

### root dataset

```{r import root data}
root = read.csv("root.txt", sep=' ', header=TRUE)
root$Tree.Number = as.factor(root$Tree.Number)

head(root)
```


```{r root stock MANOVA}
manova.root = manova(
  cbind(Girth.4.Yr, Ext.Growth.4.Yr, Girth.15.Yr, Weight.15.Yr) ~ 
    Tree.Number, 
  data=root
  )

summary(manova.root)
```
Looks like there is a significant difference between root stocks

```{r MANOVA summary}
summary.aov(manova.root)
```
Girth and weight at 15 years are the most important responses. 

*Some assumptions MANOVA makes:*

1. Samples are random. ie. root stocks aren't close or far from each other,
in more sunlight, etc.
2. Normally distributed errors for each response, and equal variances among
groups
    + This can be tested with ANOVA, t-test, etc.
    
3. Errors (residuals) follow a multivariate normal distribution
    + mvnTest is a package that tests for this
    
4. Co-variances are equal among groups


```{r mvnTest for normal dist test example}
DH.test(data = iris[iris$Species=='versicolor',1:4])

```

**What if you can't meet the assumptions required for MANOVA?**

Use **ANOSIM** to figure out if you have a problem. (Analysis of Similarity):
anosim()

*OR*

Use **PERMANOVA** (non-parametric anova): adonis2()


### darlingtonia dataset

Goal is to quantify how far apart two samples are in euclidean space. 

```{r get darlingtonia data}
darl = read.csv("DarlingtoniaData.csv", header=TRUE)
head(darl)
```
height is very large, but the other values are very small. Time to standardize
the data.

```{r z score normalization}
height = scale(darl$height, center=TRUE, scale=TRUE)
mouth_diam = scale(darl$mouth_diam, center=TRUE, scale=TRUE)
tube_diam = scale(darl$tube_diam, center=TRUE, scale=TRUE)

```

