---
title: "Final"
author: "Sean Fahey"
date: "2023-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

# About the dataset


### Reference

[Soil Dataset](https://www.kaggle.com/datasets/ramakrishnanthiyagu/soil-
compositions)

Originally distributed by LSU School of Plant, Environmental, and Soil 
Sciences (Dr. Moser Research Laboratory). 

<hr>

### Descrription

__Summary:__

Soil samples in "commerce silt loam" cotton fields in Louisiana collected 
between 1987 and 1990.

9 response variables were measured at 4 different depths (0-10cm, 10-30cm, 
30-60cm, and 60-90cm), and 3 different contours (Top, Slope, and Depression).

__Response Variables:__

- pH
- N (total Nitrogen %)
- Dens (soil density in gm/cm^3)
- P (total Phosphorous in ppm)
- Ca (total Calcium in meq/100g)
- Mg (total Magnesium in meq/100g)
- K (total Phosphorous in meq/100g)
- Na (total Sodium in meq/100g)
- Conduc (conductivity)


```{r import and clean up data}
# Import data csv file
soil = read.csv("Soils.csv")

# Correct the string entries for Depth
soil$Depth[soil$Depth == "0-10"] = "Shallow"
soil$Depth[soil$Depth == "Oct-30"] = "Medium"
soil$Depth[soil$Depth == "30-60"] = "Deep"
soil$Depth[soil$Depth == "60-90"] = "Very Deep"

# Add a group column to the front by combining Depth and Contour columns
group <- paste(soil$Depth, soil$Contour, sep="_")
soil = data.frame(Group = group, soil)

# Drop unused columns
soil = soil[,c(-2:-6)]

head(soil)
```




```{r}


```




```{r}


```




```{r}


```




```{r}


```




```{r}


```




```{r}


```




```{r}


```

















