---
title: "Project 2"
author: "Sean Fahey, Andrew Levine, Ryin Rouzbehani"
date: "2023-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(car)
library(plotly)
library(ggplot2)
```


```{r import data}
pcom.data <- read.csv("PCOMdata.csv")

# convert No to 0, yes to 1. 
pcom.data$Oligo2 = ifelse(pcom.data$Oligomenorrhea == "No Oligo",0,1)
pcom.data$Oligomenorrhea = as.factor(pcom.data$Oligomenorrhea)

str(pcom.data)
head(pcom.data)
count(pcom.data, Oligomenorrhea) # data set is unbalanced

```


```{r histograms}
#png(file="Age-histogram.png", width=400, height=250)
ggplot(
  pcom.data, 
  aes(x = Age)
  ) +
  geom_histogram(
    fill = "cornflowerblue", 
    color = "black", 
    bins=6
    ) +
  labs(title = "Age Distribution Histogram", x = "Age")
#dev.off()

#png(file="Testosterone-histogram.png", width=400, height=250)

ggplot(
  pcom.data, 
  aes(x = Testosterone.ng.mL)
  ) +
  geom_histogram(
    fill = "cornflowerblue", 
    color = "black", 
    bins=6
    ) +
  labs(title = "Testosterone Distribution Histogram", x = "Testosterone (ng/mL)")
#dev.off()

#png(file="BMI-histogram.png", width=400, height=250)
ggplot(
  pcom.data, 
  aes(x = BMI)
  ) +
  geom_histogram(
    fill = "cornflowerblue", 
    color = "black", 
    bins=6
    ) +
  labs(title = "BMI Distribution Histogram", x = "BMI")
#dev.off()

#png(file="Logit-BMI-histogram.png", width=400, height=250)
ggplot(
  pcom.data, 
  aes(x = logit(BMI))
  ) +
  geom_histogram(
    fill = "cornflowerblue", 
    color = "black", 
    bins=6
    ) +
  labs(title = "Logit BMI Distribution Histogram", x = "Logit BMI")
#dev.off()

#png(file="Oligomenorrhea-histogram.png", width=400, height=250)
ggplot(
  pcom.data, 
  aes(x = Oligo2)
  ) +
  geom_histogram(
    fill = "cornflowerblue", 
    color = "black", 
    bins=2
    ) +
  labs(title = "Oligomenorrhea Distribution Histogram", x = "Oligomenorrhea")
#dev.off()

```


```{r set up type 3 linear model}

# use a type 3 because data is unbalanced (more than 2x as many negative for oligo as positive for oligo)
options(contrasts = c("contr.sum", "contr.poly"))

# Not using the log transformed BMI because it doesn't seem to improve enough to justify it. Unfortunately we can't see if it helps residuals to transform because intrpreting residuals in a binomial glm is not really possible.

model.1 = glm(
  # assuming Testosterone, age, and BMI all interact.
  Oligo2 ~ Age * BMI * Testosterone.ng.mL,
  data=pcom.data,
  family = "binomial"
  )

options(contrasts = c("contr.treatment", "contr.poly"))

```

Type 3 model selected because we are assuming interactions between all our IV, and because the Oligomenorrhea diagnosis is imbalanced in our dataset.


```{r anova}
Anova(
  model.1, 
  type=3, 
  test.statistic="LR"
  )

```

There are no significant interactions effects on Oligomenorrhea between age, BMI, testesterone, or all three. There are also no individual effects on Oligomenorrhea from any of the three independant variables.

```{r summary}
# Anova is so bad that were not even going to show these in the presentation. Nothing here is even close to significant.
summary(model.1)
```

The summary results mirror the anova results in that they show nothing is close to statistically significant. None of the variables we tested seem to have a significant effect on the presence of Oligomenorrhea.


```{r residuals}
# cool pictures, impossible to interpret
plot(model.1)

```

"Binomial data can't be over-dispersed (supposedly), so don't worry about it. Plotting binomial residuals makes cool pictures, but interpreting them is very hard."
   - Dr. Nick Barber (fancy stats man)


```{r graph Testosterone and Age with Oligo}

#png(file="Testosterone-Age-Oligo-box.png", width=600, height=350)

plot = ggplot(
  pcom.data,
  aes(
    y = Oligo2,
    x = Age,
  )) +
  geom_boxplot(
    aes(
      x=Age, 
      group=Oligomenorrhea
      ), 
    orientation='y', 
    outlier.alpha = 0,
    varwidth = TRUE
    ) +
  geom_smooth(
    aes(x=Age, y=Oligo2),
    formula = 'y ~ x', 
    method = "glm", 
    method.args = list(family = "binomial"),
    se = FALSE,
    linetype = 2,
    color=rgb(0, 0, 0, 0.5)
    ) + 
  geom_point(
    position = position_jitter(seed=1, width=0, height=0.2),
    shape = 19,
    size = 2.3,
    color="black"
  ) +
  geom_point(
    position = position_jitter(seed=1, width=0, height=0.2),
    size = 1.7, 
    shape = 19,
    aes(color=Testosterone.ng.mL)
    ) +
  theme_grey() +
  scale_color_gradient2(
    midpoint=mean(pcom.data$Testosterone.ng.mL), 
    low="orangered", 
    high="limegreen",
    ) +
  labs(
    title = "Testosterone and Age",
    subtitle = "by Oligomenorrhea Diagnosis",
    x = "Age",
    y = "Oligomenorrhea"
  )

plot
#dev.off()

```

The regression shows a decreasing trend, but the box plots show that there is no significant difference in dispersion between the positive and negative patients. Visually, we can't confidently say there is a significant change in the slope.


```{r graph Testosterone and Age with Oligo (no box)}

#png(file="Testosterone-Age-Oligo-no-box.png", width=600, height=350)

plot = ggplot(
  pcom.data,
  aes(
    y = Oligo2,
    x = Age,
  )) +
  geom_smooth(
    aes(x=Age, y=Oligo2),
    formula = 'y ~ x', 
    method = "glm", 
    method.args = list(family = "binomial"),
    se = TRUE,
    linetype = 2,
    color=rgb(0, 0, 0, 0.5)
    ) + 
  geom_point(
    shape = 19,
    size = 2.3,
    color="black"
  ) +
  geom_point(
    size = 1.7, 
    shape = 19,
    aes(color=Testosterone.ng.mL)
    ) +
  theme_grey() +
  scale_color_gradient2(
    midpoint=mean(pcom.data$Testosterone.ng.mL), 
    low="orangered", 
    high="limegreen",
    ) +
  labs(
    title = "Testosterone and Age",
    subtitle = "by Oligomenorrhea Diagnosis",
    x = "Age",
    y = "Oligomenorrhea"
  )

plot

#dev.off()

```

This is the same plot as above, but removing the boxplots to show the line better.


```{r graph Age BMI scatter}
#png(file="BMI-Age-Oligo-scatter.png", width=600, height=350)

plot = ggplot(
  pcom.data,
  aes(
    y = BMI,
    x = Age,
    color=Oligomenorrhea
  )) +
  geom_point(
    shape = 19,
    size = 2.3,
    color="black"
  ) +
  geom_point(
    size = 1.7, 
    shape = 19,
    ) +
  geom_smooth(
    formula = 'y ~ x', 
    method = "lm", 
    se = TRUE,
    linetype = 2,
    ) + 
  theme_grey() +
  labs(
    title = "Age and BMI",
    subtitle = "by Oligomenorrhea Diagnosis",
    x = "Age",
    y = "BMI"
  )

plot

#dev.off()

```

We see that there's a possibility of a significant difference here, but it's not quite there yet. Maybe with more data this would resolve better.


```{r graph BMI Testosterone scatter}
#png(file="BMI-Testosterone-Oligo-scatter.png", width=600, height=350)

plot = ggplot(
  pcom.data,
  aes(
    y = BMI,
    x = Testosterone.ng.mL,
    color=Oligomenorrhea
  )) +
  geom_point(
    shape = 19,
    size = 2.3,
    color="black"
  ) +
  geom_point(
    size = 1.7, 
    shape = 19,
    ) +
  geom_smooth(
    formula = 'y ~ x', 
    method = "lm", 
    se = TRUE,
    linetype = 2,
    ) + 
  theme_grey() +
  labs(
    title = "Testosterone and BMI",
    subtitle = "by Oligomenorrhea Diagnosis",
    x = "Testosterone (ng/mL)",
    y = "BMI"
  )

plot

#dev.off()

```

The lines intersect which supports our interaction hypothesis, but there is no significant difference between the two groups.


```{r graph 3D scatter}
fig <- plot_ly(
  type="scatter3d",
  mode="markers",
  pcom.data, 
  x = ~Age, 
  y = ~BMI, 
  z = ~Testosterone.ng.mL, 
  color = ~Oligomenorrhea, 
  opacity = 0.8,
  colors = c('#0C4B8E', '#BF382A'),
  marker = list(
      size = 6,
      line = list(
        color = 'rgb(0, 0, 0)',
        width=1
      ))
  ) %>% layout(title="Testosterone, Age, and BMI <br>by presence of Oligomenorrhea")


fig
```

At the right angle there appears to be a slight clustering of Oligomenorrhea positive patients, but it clearly isn't showing up in our model.


```{r heatmap}

pcom.o.pos = filter(pcom.data, Oligomenorrhea == "Oligomenorrhea")
Age.pos = pcom.o.pos$Age
BMI.pos = pcom.o.pos$BMI
Testosterone.ng.mL.pos = pcom.o.pos$Testosterone.ng.mL

pcom.o.neg = filter(pcom.data, Oligomenorrhea == "No Oligo")
Age.neg = pcom.o.neg$Age
BMI.neg = pcom.o.neg$BMI
Testosterone.ng.mL.neg = pcom.o.neg$Testosterone.ng.mL


fig1 = plot_ly(
  type = 'contour', # set plot type
  x = ~Age.pos,
  z = ~BMI.pos,
  y = ~Testosterone.ng.mL.pos,
  line=list(smoothing=1.3, width=1),
  contours = list( # set contour lines to heatmap/smoothing method
    coloring = 'heatmap',
    showlabels = TRUE,
    labelfont=list(color="lightgrey")
  ),
  coloraxis="coloraxis" # select color axis so both plots share one BMI axis
  ) %>%
  layout(
    xaxis = list(title = 'Age', gridcolor = 'ffff'), # axis labels
    yaxis = list(title = 'Testosterone (ng/mL)', gridcolor = 'ffff')
    )

fig2 = plot_ly(
  type = 'contour',
  x = ~Age.neg,
  z = ~BMI.neg,
  y = ~Testosterone.ng.mL.neg,
  line=list(smoothing=1.3, width=1),
  contours = list(
    coloring = 'heatmap',  # enumerated , one of ( "fill" | "heatmap" | "lines" | "none" )
    showlabels = TRUE,
    labelfont=list(color="lightgrey")
  ),
  coloraxis="coloraxis" # matches plot 1 so BMI will share axis
  ) %>%
  layout(
    xaxis = list(title = 'Age', gridcolor = 'ffff'), 
    yaxis = list(title = 'Testosterone (ng/mL)', gridcolor = 'ffff') # --this is ignored if Y axis is shared--
    )

# build subplots
fig <- subplot(
  fig1, fig2, 
  nrows = 1, # make sure plots are side by side, not one above the other
  titleY = TRUE, 
  titleX = TRUE, 
  margin = 0.05,
  shareY = TRUE # share the y axis. (I might remove this)
  )

fig <- fig %>% # add layout options
  layout(
    title = '<b>Age and Testosterone correlation with BMI</b>',
    showlegend = FALSE, 
    coloraxis=list( #color scale and range
      colorscale="Jet", 
      cmax=35, 
      cmin=20, 
      colorbar=list(
        title=list(text="BMI")
        )
      )
    )

# add annotations for positive and negative labels on plots
annotations = list( 
  list( 
    x = 0.0775,  
    y = .956,  
    text = "Oligomenorrhea",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE,
    font = list(color= "White"),
    bgcolor="#4f4f4f"
  ),  
  list( 
    x = 0.6425,  
    y = .956,  
    text = "No Oligomenorrhea",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE,
    font = list(color= "White"),
    bgcolor="#4f4f4f"
  )
)

fig <- fig %>%layout(annotations = annotations) # add annotations to plots


fig

```

The high BMI regions between the two plots don't really overlap, which might support our findings of a cluster in the 3D scatter plot. However, this clustering is not being picked up in our model.


