---
title: 'HW 16 - NMDS'
author: "Sean Fahey"
date: "2023-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)

```

# In Class Work
__NMDS__

_similar to PCA_

- get distance matrix
- use different distance/dissimilarity methods
- algo. generates new ordination (order for values)
- compares order to initial distance matrix. Calculate goodness of fit 
(called stress)
- finds minimum stress value. Plots how samples are arranged in 2-D.

__Bray Curtis__

- doesn't vary when units change
- unaffected by addition or removal of species not found in a pair of 
communities. (if 2 sites share a set of species, they should be similar)
- unaffected by adding new communities/sites
- Can show difference in abundance, ie. if one site is far more abundance than
another even though the two sites share the same number of species.
- value of 1 = sites have nothing in common
- value of 0 = sites are identical

```{r import data}
veg = read.csv("bryceveg.txt", header=TRUE, sep=' ', row.names=1)
sites = read.csv("brycesite.txt", header=TRUE, sep=' ', row.names=1)

```


### meta-MDS
slightly fancier than normal NMDS, does refitting for us automatically

PCA gives us one principal componenet per column in dataset, MDS gives us
k components (depending on what k we specify)
```{r meta NMDS}
# k=2 means do this in 2 dimensions
bc.nmds = metaMDS(comm = veg, distance = "bray", k = 2, trace = 0)

bc.nmds
```
stress value is ~0.14. General rule of thumb is that below 0.2, below 0.1 is 
very good, 0 is best.

```{r plot ordination distances}
pl = ordiplot(
  bc.nmds, 
  type='none', 
  main=paste0( # plot title
    'Stress = ', 
    round(bc.nmds$stress,dig=3)
    )
  )

points(pl, what='sites', col="black", pch=4) # add points for the sites

points(pl, what='species', col='tomato', pch=20) # add points for the species

legend( # add legend at bottom
  'bottomright', 
  legend=c('Sites', 'Species'), 
  col=c('black', 'tomato'), 
  pch=c(4,20)
  )
```

### Stress

Stress is the goodness of fit measurement. A stress of 0.2 means roughly 80%
of the variation in the original distance matrix can be explained by the 2
MDS components. 

```{r stress plot}


```

### Ordination Plots

```{r fancy ordination plot}
pl = ordiplot(
  bc.nmds, 
  type='none', 
  main=paste0( # plot title
    'Stress = ', 
    round(bc.nmds$stress,dig=3)
    )
  )

# add points for sites where berries present
points(pl, veg$berrep>0, what='sites', col="tomato", pch=20)

# add points for sites where berries absent
points(pl, veg$berrep == 0, what='sites', col='black', pch=20)

legend( # add legend at bottom
  'bottomright', 
  legend=c('Berries present', 'Berries absent'), 
  col=c('tomato', 'black'), 
  pch=c(20,20),
  title=expression(italic(Berries ~ repens))
  )

```


```{r add convex hull}
pl = ordiplot(
  bc.nmds, 
  type='none', 
  main=paste0( # plot title
    'Stress = ', 
    round(bc.nmds$stress,dig=3)
    )
  )

# add points for sites where berries present
points(pl, veg$berrep>0, what='sites', col="tomato", pch=20)

# add points for sites where berries absent
points(pl, veg$berrep == 0, what='sites', col='black', pch=20)

legend( # add legend at bottom
  'bottomright', 
  legend=c('Berries present', 'Berries absent'), 
  col=c('tomato', 'black'), 
  pch=c(20,20),
  title=expression(italic(Berries ~ repens))
  )

# add an outline where the groups are found.
ordiellipse(
  bc.nmds, 
  kind='ehull', # how to include points, this means all (?)
  draw="polygon", 
  groups=veg$berrep>0, 
  col=c('black', 'red'), 
  label=F
  )

```


```{r adding outside info}
# make the depth column a factor
sites$depth = as.factor(sites$depth)

pl = ordiplot(
  bc.nmds, 
  type='none', 
  main=paste0( # plot title
    'Stress = ', 
    round(bc.nmds$stress,dig=3)
    ),
  las = 1
  )

# add points colored by sites column
# this gives me color number 5 and 6, change the addition to change the color 
# pair
points(pl, what='sites', col=as.numeric(sites$depth)+5, pch=20)

legend( # add legend at bottom
  'bottomright', 
  legend=levels(sites$depth),
  col=6:7, # make the colors match the points
  pch=20,
  cex=1.2,
  title="Depth"
  )

```


```{r adding contours}
pl = ordiplot(
  bc.nmds, 
  type='none', 
  main=paste0( # plot title
    'Stress = ', 
    round(bc.nmds$stress,dig=3)
    ),
  las = 1
  )

# add points colored by sites column
# this gives me color number 5 and 6, change the addition to change the color 
# pair
points(pl, what='sites', col='black', pch=4)

legend( # add legend at bottom
  'bottomright', 
  legend="Sites",
  col='black', # make the colors match the points
  pch=4,
  )

# add contour markers for elevation of site
ordisurf(bc.nmds, sites$elev, main='', add=TRUE)
```


```{r envfit}
ef2 = envfit(bc.nmds, sites[,c('elev', 'av', 'slope')])
ef2
```


```{r envfit ordination plot}
ordiplot(
  bc.nmds, 
  main=paste0(
    "Stress = ", 
    round(bc.nmds$stress, dig=3)), 
  las=1
  )

plot(ef2)

```
There's no units on an NMDS plot, because we converted to a rank-order. 
Typically there's not even ticks on an axis when publishing the values. 

----

### Standardization and transformation

- metaNMDS applies a default transformation
- some common transformations:
  1. sq root transformation
  2. 4th root (double sq root)
  3. log-transform
  4. Wisconsin


# Homework

_instructions_

- Run a NMDS on the bird data (macnally_full2.csv, leave out columns 1&2, 
SITE + HABITAT), using Bray-Curtis as the distance metric with the vegan 
metaMDS function

- Generate a NMDS ordination plot showing the sites only (no species arrows), 
and draw ellipsoid hulls around each of the 6 habitat types to show how 
habitat maps onto the ordination.

  - Color the site points by HABITAT type (column 2 of data frame)

  - Use the ordiellipse() function, and the groups argument should be defined 
using the HABITAT data in column 2 of the data frame

  - Include a legend in the plot to show which point color maps to which 
habitat type

- Based on your plot, describe which (if any) of the habitats differ in 
community composition



### NMDS
```{r import dataset}
bird = read.csv("macnally_full2.csv", header=TRUE, sep=',', row.names=1)
bird$HABITAT = as.factor(bird$HABITAT)

```

NMDS summary
```{r NMDS}
bird.nmds = metaMDS(comm = bird[c(-1,-2)], distance = "bray", k = 2, trace = 0)

bird.nmds
```

ordination plot with just sites
```{r ordination plot}
bird.plot = ordiplot(
  bird.nmds, 
  type='none', 
  main=paste0( # plot title
    'Stress = ', 
    round(bird.nmds$stress,dig=3) # add stress label to title
    ),
  xaxt='none', # remove tick marks from NMDS plots 
  yaxt='none'
  ) 

# add data points
points(
  bird.plot, 
  what='sites', 
  col=as.numeric(bird$HABITAT), # color by habitat
  pch=as.numeric(bird$HABITAT) # shape by habitat
  )

legend( # add legend at bottom left
  'bottomleft', 
  legend=levels(bird$HABITAT),
  col=1:6, # make the colors match the points
  pch=1:6, # make the shapes match
  cex=0.9,
  title="Depth"
  )

ordiellipse(
  bird.nmds, 
  kind='ehull', # how to include points, this means all (?)
  draw="polygon", 
  col=1:6,
  alpha=0.2,
  groups=bird$HABITAT, 
  label=F
  )

```








